---
title: "UMN Extension Carbon Challenge calculations - 2021"
author: "Matt Russell"
date: "6/4/2021"
output:
  word_document: default
  html_document: default
  
always_allow_html: yes
---

The following is a document to perform calculations for the University of Minnesota Extension's Tree and Woodland Carbon Capture Challenge. The general approach is to (1) input tree and woodland diameter measurements, (2) calculate the aboveground biomass of each tree, and (3) calculate the carbon content in the tree or woodland. 

```{r, echo = F,  message=FALSE, warning=F}
library(tidyverse)
library(googlesheets4)
library(knitr)
library(formattable)
library(kableExtra)
options(kableExtra.auto_format = FALSE)
```

Here's some test data explained in the [Google Doc instructions](https://docs.google.com/document/d/1-4J-NvyEF7Y-0F6aeuXYBtGkY5y8Cu7sZ0k9lvjON4o/edit#heading=h.n8brrdl98b58).

Example for a single tree:

```{r, echo = F}
tree <- tribble(
  ~SPGRP_JENKINS, ~TIME, ~DIA,
  "Maple/oak/hickory/beech", "BEFORE", 10.5,
  "Maple/oak/hickory/beech", "AFTER", 11.2
)
```

The Jenkins et al. 2003 paper describes a set of species-specific equations to estimate aboveground biomass in trees. The calculation relies on species and diameter at breast height (inches) and calculates biomass in pounds: [http://www.fs.fed.us/ne/durham/4104/papers/Heathbiomass_eqns.pdf](http://www.fs.fed.us/ne/durham/4104/papers/Heathbiomass_eqns.pdf)  

```{r, echo = F}
# Here is the function:

total_AGB<-function(SPGRP_JENKINS,DIA){
if(SPGRP_JENKINS=='Aspen')
  {JENKINS_TOTAL_B1=-2.2094; JENKINS_TOTAL_B2=2.3867}
if(SPGRP_JENKINS=='Cedar/larch')	
  {JENKINS_TOTAL_B1=-2.0336; JENKINS_TOTAL_B2=2.2592}
if(SPGRP_JENKINS=='Maple/oak/hickory/beech')	
  {JENKINS_TOTAL_B1=-2.0127; JENKINS_TOTAL_B2=2.4342}
if(SPGRP_JENKINS=='Mixed hardwood')	
  {JENKINS_TOTAL_B1=-2.48; JENKINS_TOTAL_B2=2.4835}
if(SPGRP_JENKINS=='Pine')	
  {JENKINS_TOTAL_B1=-2.5356; JENKINS_TOTAL_B2=2.4349}
if(SPGRP_JENKINS=='Soft maple/birch')	
  {JENKINS_TOTAL_B1=-1.9123; JENKINS_TOTAL_B2=2.3651}
if(SPGRP_JENKINS=='Spruce')	
  {JENKINS_TOTAL_B1=-2.0773; JENKINS_TOTAL_B2=2.3323}
if(SPGRP_JENKINS=='True fir/hemlock') 	
  {JENKINS_TOTAL_B1=-2.5384; JENKINS_TOTAL_B2=2.4814}
  # Calculate total aboveground biomass of each tree (AGB)
  # Note the equation predicts biomass in kg, but we  convert it into pounds 
  AGB=(exp(JENKINS_TOTAL_B1+JENKINS_TOTAL_B2*log(DIA*2.54)))*2.20462
    return(AGB=AGB)}
```

Now, we can apply the function total_AGB to estimate total aboveground biomass and add it as a column in our tree data set. Then, we'll calculate the carbon content, assuming half of biomass is carbon: 

```{r, echo = F}
tree$AGB<-mapply(total_AGB,SPGRP_JENKINS=tree$SPGRP_JENKINS,DIA=tree$DIA)
tree$C<-tree$AGB*0.5

# Calculate CO2 equivalents
tree$CO2e <- tree$C * 3.667

tree
```

## Load the pre-growing season measurements of the Carbon Capture Challenge

```{r, echo = F}
tree2021 <- read_sheet("https://docs.google.com/spreadsheets/d/1sjRe8pi08C4jY8TlDBeRKxmhqMlo85eB5wmqNovsiOU/edit?usp=sharing")

tree2021 <- tree2021 %>% 
  filter(Year == 2021)
tree2021$AGB1<-mapply(total_AGB,SPGRP_JENKINS=tree2021$Spp2,DIA=tree2021$DBH1)
tree2021$DBH1<-round(tree2021$DBH1, 1)
tree2021$C1<-round(tree2021$AGB1*0.5, 0)

tree2021$AGB2<-mapply(total_AGB,SPGRP_JENKINS=tree2021$Spp2,DIA=tree2021$DBH2)
tree2021$DBH2<-round(tree2021$DBH2, 1)
tree2021$C2<-round(tree2021$AGB2*0.5, 0)
tree2021$dC<-tree2021$C2-tree2021$C1


# Calculate CO2 equivalents
tree2021$CO2e1 <- tree2021$C1 * 3.667
tree2021$CO2e2 <- tree2021$C2 * 3.667
tree2021$dCO2e<-tree2021$CO2e2-tree2021$CO2e1

 tree2021 <- tree2021 %>% 
      rename(`Team name` = Team,
         Species = Spp,
           `Tree diameter, time 1 (inches)` = DBH1 ,
           `Carbon in tree, time 1 (pounds)` = C1,
         `Tree diameter, time 2 (inches)` = DBH2 ,
           `Carbon in tree, time 2 (pounds)` = C2,
         `Carbon sequestered (pounds)` = dC)

 tree_show <- tree2021[, c(3,6,10,12,14,16,17)] %>% 
  head(40)
 tree_show
tree_show %>% 
       arrange(-`Carbon sequestered (pounds)`) %>% 
   kable("html", caption = 'Entries in the MN Carbon Capture Challenge (Single tree category).') %>%
  kable_styling()


tree2021_2 <- tree2021 %>% 
      rename(`Tree diameter  (inches)` = `Tree diameter, time 2 (inches)`,
           `Carbon stored in tree (pounds)` = `Carbon in tree, time 2 (pounds)`,
         `Carbon sequestered this year (pounds)` = `Carbon sequestered (pounds)`)

tree_show2 <- tree2021_2[, c(3,6,10,12,14,16,17)] %>% 
  head(40)
tree_show2
 
tree_show2 %>% 
       arrange(-`Carbon sequestered this year (pounds)`) %>% 
   kable("html", caption = 'Entries in the MN Carbon Capture Challenge (Single tree category).') %>%
  kable_styling()


p.tree1 <- ggplot(tree2021, aes(x = reorder(`Team name`, `Carbon in tree, time 1 (pounds)`), y = `Carbon in tree, time 1 (pounds)`)) +
  geom_bar(stat =  "identity", fill="steelblue", col = "black") +
  coord_flip() +
  scale_y_continuous(limits = c(0,26000))+
  labs(x = "Team name",
       y = "Carbon stored in tree (pounds)") +
    theme(panel.background = element_rect(fill = "NA"),
        axis.line = element_line(color = "black"))
p.tree1

 ggsave(filename =  "C:\\Users\\russellm\\Documents\\Projects\\CarbonChallenge\\2021\\p.tree.png",
        plot=p.tree1, width=6, height=4,units="in",scale=1)

 tree2021 <- tree2021 %>% 
     filter(!is.na(`Tree diameter, time 2 (inches)`)) 
   
tree2021_long <- tree2021 %>% 
  select(`Team name`, `Carbon in tree, time 1 (pounds)`, `Carbon in tree, time 2 (pounds)`) %>% 
  pivot_longer(!`Team name`, names_to = "Time", values_to = "Carbon")
  
Seasons <- c("Beginning of growing season \n(May 2021)","End of growing season \n(September 2021)")

p.dC<- ggplot(tree2021_long, aes(x = Time, y = Carbon, group = `Team name`)) +
  geom_point(col="steelblue", size = 2) +
  geom_line()+
  scale_x_discrete(labels= Seasons) +
  labs(x = " ",
       y = "Carbon stored in tree (pounds)",
       title = "Single Tree Category") +
    theme(panel.background = element_rect(fill = "NA"),
        axis.line = element_line(color = "black"))
p.dC

ggsave(filename =  "C:\\Users\\russellm\\Documents\\Projects\\CarbonChallenge\\2021\\p.dc.png",
       plot=p.dC, width=4, height=6,units="in",scale=1)
```

Here is some example data from a woodland:

```{r, echo = F}
wood <- tribble(
  ~SPGRP_JENKINS, ~TIME, ~DIA,
  "Maple/oak/hickory/beech", "BEFORE", 6.7,
  "Maple/oak/hickory/beech", "BEFORE", 7.7,
  "Pine", "BEFORE", 8.9,
  "Maple/oak/hickory/beech", "BEFORE", 5.9,
  "Maple/oak/hickory/beech", "BEFORE", 8.9,
  "Pine", "BEFORE", 10.3,
  "Pine", "BEFORE", 11.5,
  "Maple/oak/hickory/beech", "BEFORE", 8.8,
  "Maple/oak/hickory/beech", "AFTER", 7.0,
  "Maple/oak/hickory/beech", "AFTER", 7.8,
  "Pine", "AFTER", 9.5,
  "Maple/oak/hickory/beech", "AFTER", 6.2,
  "Maple/oak/hickory/beech", "AFTER", 9.4,
  "Pine", "AFTER", 10.3,
  "Pine", "AFTER", 11.7,
  "Maple/oak/hickory/beech", "AFTER", 8.9
)
```

Remember, participants are measuring 1/10th acre plots. We'll expand to this value in the calculation.

```{r, echo = F}
PLOT_SIZE = 10

wood$AGB<-mapply(total_AGB,SPGRP_JENKINS=wood$SPGRP_JENKINS,DIA=wood$DIA)
wood$C<-wood$AGB*0.5

wood$C_sum<-wood$C*PLOT_SIZE

wood_grouped <- group_by(wood, TIME)

plot.C <- summarize(wood_grouped, 
  CPA = sum(C_sum))
plot.C$CO2e <- plot.C$CPA * 3.667
plot.C
```


```{r, echo = F}
wood2021 <- read_sheet("https://docs.google.com/spreadsheets/d/1ekdnxAnmO0PfVAmA-aZ1j6lNdwNKMGz7WLK1vBttmtA/edit?usp=sharing")
PLOT_SIZE = 10


wood2021 <- wood2021 %>% 
  filter(Year == 2021)

wood2021$AGB1<-mapply(total_AGB, SPGRP_JENKINS=wood2021$Spp2, DIA=wood2021$DBH1)
wood2021$C1<-wood2021$AGB1*0.5
wood2021$C_sum1<-wood2021$C1*PLOT_SIZE

wood2021$AGB2<-mapply(total_AGB, SPGRP_JENKINS=wood2021$Spp2, DIA=wood2021$DBH2)
wood2021$C2<-wood2021$AGB2*0.5
wood2021$C_sum2<-wood2021$C2*PLOT_SIZE

plot.C <- wood2021 %>% 
  group_by(Team) %>% 
  summarize(CPA1 = sum(C_sum1),
            CPA2 = sum(C_sum2))

plot.C$CO2e1 <- plot.C$CPA1 * 3.667
plot.C$tons1 <- plot.C$CPA1 / 2000

plot.C$CO2e2 <- plot.C$CPA2 * 3.667
plot.C$tons2 <- plot.C$CPA2 / 2000

plot.C

`%notin%` <- Negate(`%in%`)
plot.C_long <- plot.C %>% 
  pivot_longer(!`Team`, names_to = "Time", values_to = "Carbon") %>% 
  filter(Time %in% c("tons1", "tons2"),
         Team %notin% c("Sleeping Bear Tree Farm"))
 
wood_show <- plot.C[, c(1, 5, 7)] 

wood_show <- wood_show %>% 
      rename(`Team name` = Team,
           `Carbon in woods, time 1 (tons per acre)` = tons1,
           `Carbon in woods, time 2 (tons per acre)` = tons2) %>% 
  mutate(`Carbon sequestered (tons per acre)`= `Carbon in woods, time 2 (tons per acre)`-`Carbon in woods, time 1 (tons per acre)`)

wood_show %>% 
       filter(`Team name` %notin% c("Sleeping Bear Tree Farm")) %>% 
       arrange(-`Carbon sequestered (tons per acre)`) %>% 
   kable("html", caption = 'Entries in the MN Carbon Capture Challenge (Woodland category).') %>%
  kable_styling()


p.wood <- ggplot(plot.C, aes(x = reorder(Team, tons1), y = tons1)) +
  geom_bar(stat =  "identity", fill="darkgreen", col = "black") +
  coord_flip() +
  scale_y_continuous(limits = c(0,80))+
  labs(x = "Team name",
       y = "Carbon stored in woodland (tons per acre)") +
    theme(panel.background = element_rect(fill = "NA"),
        axis.line = element_line(color = "black"))
p.wood

 ggsave(filename =  "C:\\Users\\russellm\\Documents\\Projects\\CarbonChallenge\\2021\\p.wood.png",
        plot=p.wood, width=6, height=4,units="in",scale=1)

Seasons <- c("Beginning of growing season \n(May 2020)","End of growing season \n(September 2020)")

p.dC.plot<- ggplot(plot.C_long, aes(x = Time, y = Carbon, group = `Team`)) +
  geom_point(col="darkgreen", size = 2) +
  geom_line()+
  scale_x_discrete(labels= Seasons) +
  labs(x = " ",
       y = "Carbon stored in woodland (tons per acre)",
       title = "Woodland Category") +
    theme(panel.background = element_rect(fill = "NA"),
        axis.line = element_line(color = "black"))
p.dC.plot

ggsave(filename =  "C:\\Users\\russellm\\Documents\\Projects\\CarbonChallenge\\p.dc.plot.png",
       plot=p.dC.plot, width=4, height=6,units="in",scale=1)
```
